Synthesis of Linear Functional Programs

### 2021-07-22 article/article.tex
optimization, recursion extra

\mypara{Recursion} The main idea behind synthesis of recursive
programs is the labeling of the main goal and the addition of its
type, under that name, to the unrestricted context. That is, to
synthesize a recursive function of type $A \lolli B$ named \emph{f},
the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, f{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($f{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.
%
However, we must restrict immediate uses of the recursive call since
otherwise every goal would have a trivial proof (a non-terminating
function that just calls itself), shadowing relevant solutions.
Instead, our framework allows the use of recursion only after having
deconstructed a recursive ADT via the following invariant: the
recursive hypothesis can only be used in \emph{recursive branches of
  ADT deconstruction}, i.e. the recursive call should only take
``smaller'', recursive, hypothesis as arguments. To illustrate, in any
recursive function with a list argument (whose type is defined as
\mintinline{haskell}{data List = Nil | Cons (A * List)}), recursive
calls are only allowed when considering a judgment of the form
$\textrm {List} \vdash C$, i.e.~when a list value is available to
produce the goal $C$, and only in the \emph{Cons}
branch. % (TODO: rewrite? O bold é meio estranho não?)

@@

\mypara{Refinement Types} Refinement types are defined as a types
with a name and with a predicate (a non-existing predicate is the same as it
being always \emph{true}); dependent types are functions with
refinement types in which the argument type name can be used in the predicates
of the return type (e.g. $x\{\mathsf{Int}\} \lolli y\{\mathsf{Int}\ \vert\ y = x\}$ specifies a function that
takes an Int named $x$ and produces an Int whose value is equal to the input $x$). We extend
the types syntax with our refinement types syntax:
\[
\begin{tabular}{lclc}
    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle \{\tau\ \vert\ P\}$ \\
    $P$ & $\ ::=\ $ & $P = P\ \vert\ P \neq P\ \vert\ P \vee P\ \vert\ P \wedge
    P\ \vert\ P \Rightarrow P\ \vert\ n = n\ \vert\ n \neq n$ \\
    & & $\vert\ n \leq n\ \vert\ n \geq n\ \vert\ n < n\ \vert\ n > n\ \vert\
    true\ \vert\ false\ \vert\ \langle\emph{name}\rangle$ \\
    $n$ & $\ ::=\ $ & $n * n\ \vert\ n + n\ \vert\ n - n\ \vert\
    \langle\emph{natural}\rangle\ \vert\ \langle\emph{name}\rangle$
    \\
\end{tabular}
\]
The addition of refinement types to the synthesizer doesn't
interfere with the rest of the process. We define the following right and
left rule, to synthesize or consume in synthesis a refinement type, where
$\textsc{getModel}(p)$ is a call to an SMT solver that returns a model of an
uninterpreted function that satisfies
$\forall_{\tilde{1},\tilde{2},\dots,\tilde{n}}\ h_{\tilde{1}} \Rightarrow h_{\tilde{2}} \Rightarrow \dots
\Rightarrow h_{\tilde{n}} \Rightarrow p$, where $\tilde{i}$ is the refinement type
name and $h_{\tilde{i}}$ its predicate for every refinement type in the
propositional contexts; and $\textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}})$ is a call to an SMT solver that
determines universal satisfiability of the implication between predicates (the
left focused proposition subtypes the goal).

\begin{mathpar}
    \infer*[right=(refR)]
    { \textsc{getModel}(p) = M }
    {\Theta/\Theta';\Rho;\Gamma;\Delta \vdash M : \tilde{x}\{A\ \vert\
    p\}\Uparrow }
    \and
    \infer*[right=(refL)]
    { \textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}}) }
    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\tilde{y}\{A\ \vert\ p_{\tilde{y}}\}\Downarrow\ \vdash x : \tilde{x}\{A\ \vert\ p_{\tilde{x}}\} }
\end{mathpar}

>>>>

\mypara{Recursion} The main idea behind synthesis of recursive
programs is the labeling of the main goal and the addition of its
type, under that name, to the unrestricted context. That is, to
synthesize a recursive function of type $A \lolli B$ named \emph{f},
the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, f{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($f{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.
%
However, we must restrict immediate uses of the recursive call since
otherwise every goal would have a trivial proof (a non-terminating
function that just calls itself), shadowing relevant solutions.
Instead, our framework allows the use of recursion only after having
deconstructed a recursive ADT via the following invariant: the
recursive hypothesis can only be used in \emph{recursive branches of
  ADT deconstruction}, i.e. the recursive call should only take
``smaller'', recursive, hypothesis as arguments. To illustrate, in any
recursive function with a list argument (whose type is defined as
\mintinline{haskell}{data List = Nil | Cons (A * List)}), recursive
calls are only allowed when considering a judgment of the form
$\textrm {List} \vdash C$, i.e.~when a list value is available to
produce the goal $C$, and only in the \emph{Cons}
branch. Furthermore, we also forbid the usage of the recursive function when
synthesizing arguments to use it.
% TODO : não sei se precisa de melhor explicação mas foi uma coisa que fiz para
% não gerar um programa recursivo.
% (TODO: rewrite? O bold é meio estranho não?)

@@

\mypara{Refinement Types} Refinement types are defined as a types
with a name and with a predicate (a non-existing predicate is the same as it
being always \emph{true}); dependent types are functions with
refinement types in which the argument type name can be used in the predicates
of the return type (e.g. $x\{\mathsf{Int}\} \lolli y\{\mathsf{Int}\ \vert\ y = x\}$ specifies a function that
takes an Int named $x$ and produces an Int whose value is equal to the input $x$). We extend
the types syntax with our refinement types syntax:
\[
\begin{tabular}{lclc}
    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle \{\tau\ \vert\ P\}$ \\
    $P$ & $\ ::=\ $ & $P = P\ \vert\ P \neq P\ \vert\ P \vee P\ \vert\ P \wedge
    P\ \vert\ P \Rightarrow P\ \vert\ n = n\ \vert\ n \neq n$ \\
    & & $\vert\ n \leq n\ \vert\ n \geq n\ \vert\ n < n\ \vert\ n > n\ \vert\
    true\ \vert\ false\ \vert\ \langle\emph{name}\rangle$ \\
    $n$ & $\ ::=\ $ & $n * n\ \vert\ n + n\ \vert\ n - n\ \vert\
    \langle\emph{natural}\rangle\ \vert\ \langle\emph{name}\rangle$
    \\
\end{tabular}
\]
The addition of refinement types to the synthesizer doesn't
interfere with the rest of the process. We define the following right and
left rule, to synthesize or consume in synthesis a refinement type, where
$\textsc{getModel}(p)$ is a call to an SMT solver that returns a model of an
uninterpreted function that satisfies
$\forall_{\tilde{1},\tilde{2},\dots,\tilde{n}}\ h_{\tilde{1}} \Rightarrow h_{\tilde{2}} \Rightarrow \dots
\Rightarrow h_{\tilde{n}} \Rightarrow p$, where $\tilde{i}$ is the refinement type
name and $h_{\tilde{i}}$ its predicate for every refinement type in the
propositional contexts; and $\textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}})$ is a call to an SMT solver that
determines universal satisfiability of the implication between predicates (the
left focused proposition subtypes the goal).
\begin{mathpar}
    \infer*[right=(refR)]
    { \textsc{getModel}(p) = M }
    {\Theta/\Theta';\Rho;\Gamma;\Delta \vdash M : \tilde{x}\{A\ \vert\
    p\}\Uparrow }
    \and
    \infer*[right=(refL)]
    { \textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}}) }
    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\tilde{y}\{A\ \vert\ p_{\tilde{y}}\}\Downarrow\ \vdash x : \tilde{x}\{A\ \vert\ p_{\tilde{x}}\} }
\end{mathpar}

\mypara{Optimizations} To speed up the process and get a cleaner and sometimes more correct
output, we add a rule that lets us ``skip some rules'' if left focused on a
$\bang$-ed proposition, and the goal is $\bang$-ed:
\begin{mathpar}
    \infer*[right=($\bang\Downarrow$L)]
    { \Theta/\Theta';\Rho;\Gamma;\Delta; A\Downarrow\ \vdash M : C}
    {\Theta/\Theta';\Rho;\Gamma;\Delta; \bang A\Downarrow\ \vdash M : \bang C}
\end{mathpar}
